\documentclass[12pt]{report}

\include{header}

\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subfigure}
\usepackage{tikz}

\usepackage{listings}
\lstset{language=Java}

\newcommand{\Ans}{\hfill {\Huge\begin{tabular}{|c|} \hline \hspace{2.5cm} \hfill \\ \hline \end{tabular}} }

\newcommand{\AnsBox}[1]{
\hspace{-0.6cm}\fbox{
\begin{minipage}[b][#1][c]{17.6cm}
\hfill
\end{minipage}}
}

\begin{document}

%%%TITLE PAGE%%%
{\large
\begin{center}
\textsc{The University of Western Ontario}
\end{center}
\begin{center}
\textbf{\textsc{Computer Science 1026b, Spring 2010}}\\
\textbf{\textsc{Computer Science Fundamentals I}}\\
{\large{Final Exam} }\\
\textsc{2:00pm - 5:00pm, Friday, April 16, 2010.}
\end{center}

\begin{center}
\begin{tabular}{ r c}
\textsc{Name (last, first):} & \rule{2.5in}{0.01in} \\
\textsc{Student Number: }& \rule{2.5in}{0.01in}
\end{tabular}
\end{center}

{\normalsize
\vspace{20pt}
\begin{center}\textbf{Instructions}\end{center}
\begin{enumerate}
\item TURN OFF YOUR CELL PHONE AND ALL ELECTRONIC DEVICES.
\item Do not turn this page until instructed to do so.
\item Cheating (e.g. copying someone else's solutions, using prohibited material like your textbook and notes, 
or using electrical communication devices) will result in a zero grade and possible expulsion from the course.We reserve the right to orally examine you regarding your solutions and then regrade accordingly.
\item For each question, provide your solution in the answer box. {\bf All writing outside boxes will be ignored} by the markers. 
\item The exam document consists of twenty-three (23) pages, the last five (5) of them
       are meant for rough work and will not be marked.No other paper document than the exam document
 is permitted during the exam.
\item {\bf The entire exam document}, including the five (5) pages of rough work,
 {\bf must be returned} to 
      the proctoring personnel before leaving the exam room.
     Failing to do will result in a zero grade.
\item Only those exams written in pen are eligible for regrading.
\item Correct java code is preferable but pseudo-code 
(i.e. something resembling code) is 
acceptable and if appropriate will be accepted for marks.
\end{enumerate}
\vspace{10pt}
}

\begin{center}
{\Large
\begin{tabular}{||c|c|r||} \hline 
  \S 1 & 20 &\hspace{10mm} \hfill \\ \hline 
  \S 2 & 20 &       \\ \hline
  \S 3 & 40 &      \\ \hline
  \S 4 &  20 &    \\ \hline
Total & 100 & \\ \hline 
\end{tabular}
}
\end{center}
}
\clearpage

\begin{titlepage}
\newpage
\mbox{}
\newpage
\end{titlepage}
%%%END TITLE PAGE%%%

\input{VeryShort.tex}
\input{Short.tex}
\input{Long.tex}
\input{VeryLong.tex}

%\input{RoughPage.tex}

\end{document}
